On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification

### 2020-10-15 paper/cav_determinisim_main.tex
Pushing changes from discussion with Greg.

\documentclass[letterpaper, 10 pt, journal, twoside]{IEEEtran}
% \input{archive/main.config.tex}
\usepackage[maxnames=6,firstinits=true,doi=false,url=false,isbn=false]{biblatex}
% \addbibresource{IROS.bib}
\addbibresource{cav_determinism.bib}
\usepackage{array}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{threeparttable}
\usepackage{balance} %balance columns
\hyphenation{op-tical net-works semi-conduc-tor}
\usepackage{enumitem}
\usepackage{hyperref} % Add  [ocgcolorlinks,pdfusetitle] before hyperref for colored links
% \usepackage{subfig}
\usepackage{subcaption}
\usepackage{dblfloatfix}
\usepackage{booktabs}
\usepackage{authblk}
\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
% \pagenumbering{gobble} %suppresses page numbers
\usepackage{balance} %balance columns
\usepackage{todonotes}
\usepackage{siunitx}
% \usepackage{draftwatermark}

@@

\begin{abstract}
\noindent 
Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
%
Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
\todo[inline,color=green!40]{GC Alternative: Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
%
% KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
% These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
%
However, game engines do not need to be deterministic per se. % when used for games. 
%
This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation precision} for specific configurations and utilisation of the software and hardware, respectively. 
%
A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.

\todo[inline,color=green!40]{GC Alternative: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance. }
\todo[inline]{Review again at end. Potentially give away more on results to motivate the reader to continue.}

@@

% Game engines for SBV
Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation. 
%
However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include determinism.
%
% What is a deterministic system
In this context, deterministic refers to the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
% 
% Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions. Specifically, we focus on the variance of actor paths within a given scenario. 
%
Herein we adopt the engineering terms of \textit{precision} and \textit{tolerance} to describe how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions. %
%
As an analogy, the simulator can be thought of as manufacturing a product, producing data traces and if tasked to repeatedly make the same product we must ask how \textit{precise} can it be and what will the \textit{tolerance} be when comparing any differences between products. The precision of any computer based task is casually assumed to be absolute, but technically this is equivalent to the representation of numbers within it, e.g. 64-bit floating point \textbf{[ref]}. %
%
Real manufacturing can only aspire to achieve such precision and hence the need for tolerances on design specifications to account for the variance in real-world processes. If a simulator is deterministic then it will produce results with some degree of floating point precision and zero tolerance, i.e. with zero variance on any measure of distinction between simulation output traces. 
%
This is not to be confused with \textit{accuracy} which describes the extent to which the simulation complies to a known standard, in many cases a simulation is held to the standard of the real world, where any divergence from this standard is termed the \textit{reality gap}.

@@

% Verification and Determinism 
% Why is determinism important for verification
%For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
% KE Text
Determinism is a key prerequisite for simulation during AV development and testing. A deterministic simulation environment guarantees that tests are repeatable. Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated. This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.

@@

% Challenges of SBV
When used for gaming, game engines do not need to be deterministic nor do they even have any requirements on the limits of permissible tolerance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers. It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience. However, the situation is very different for AV development and testing. Thus, our main research questions is:
How can one assess the extent to which a simulation environment is deterministic? 
\todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}

In pursuit of an answer to this question we have investigated how non-determinism of a game engine, namely Unreal Engine that powers CARLA, when used for simulation-based AV verification affects the simulation results. 
%
In our case study, scenarios between pedestrian and vehicle actors have been analysed to identify conditions that result in non-deterministic simulation traces by analysis of actor position variance.
%
By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.

\todo[inline,color=green!40]{GC Addition: By analysing actor position variance we find that the Carla simulator is not deterministic under certain conditions. However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}

The insights gained from this case study have motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment. 
%
Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have. In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results. 
%
\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have. In particular, they can gain a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this. }
%
% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
%We present a case study to illustrate the use of this method. The study highlights the effects of non-determinism of a game engine on simulation results. Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
%
%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
%

@@

Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.  %However, when these engines are used for AV development then deterministic behaviour is required.

\todo[inline]{GC line added to introduction. We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}. Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion. But it is important that we demonstrate knowledge of others working on this in other areas. {\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}

\todo[inline,color=green!40]{GC Addition: Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}. }

This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
%
% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in AV simulations. Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.

@@

% This process allows hardware scalability, but the rendering will become of jerky quality on slower machines. These engines update at fixed intervals and render whenever they can, which is not steady. This results in what is so called residual lag\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, where the engine is trying to render between two consecutive updates. In this case, the engine will use extrapolation techniques to give an estimate of where the object should be. This often is sufficient for gaming purposes and unnoticeable to the user, in fact it improves the stuttery motion.

Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences. Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy. In contrast, simulation accuracy is essential in AV development and testing, especially for perception stack testing, but also for visual inspection of accidents or near misses. If necessary, it is acceptable to achieve the required level of accuracy at the cost of real-time performance. For example, for perception stack testing the sensors need to get input that is as close as possible to what the real input would be, i.e.\ with no or minimal extrapolation or lag. This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.

\todo[inline,color=green!40]{GC Edit: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences. Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision. The gamer neither wishes a faithful representation of reality, i.e. can be low accuracy, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e. can be low precision. In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses. If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance. For example, for perception stack testing the sensors need to get input that repeatable so that if any software bugs are found they can be re-played and the issue resolved. This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.}

\todo[inline]{KIE: ... by slowing down the simulation to enable more extensive physics calculations?}

@@

Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
Thus, the results of arithmetic operations must fit into the given bit width, which is achieved through rounding to the nearest representable value. This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}. %
%
% SEE BELOW \todo[inline]{Check the points on floating point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
\todo[inline,color=green!40]{GC Addition (added reference on non-associative fp addition Kapre2007, Whitehead2011, empirical-analysis-of-flaky-tests, see para below:)

Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour. Calculation results may differ due to the nature of non-associative floating point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}. Some authors even suggest avoiding the use of floating point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}. 
%
Here we are assuming these calculations are performed on a single core and therefore concurrency of multiple threads is not an issue which is discussed in the section below titled Scheduling, Concurrency and Parallelisation.}

%
% Floating-point representation of real numbers is limited with fixed register sizes resulting in errors of rounding where loss of bits occur which are rounded or truncated and cancellation which happens where two similar values are subtracted.

@@

However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events. This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
%
\todo[inline,color=green!40]{GC addition: Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}. }
%
Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.

@@

\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}. }

@@

To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.

@@

We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.
\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
%
\todo[inline]{Complete the section structure as this is a very long section.}

@@

This case study draws on a setup used to verify an urban last-mile mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
%
Thus, the scene for our investigation is the T-junction shown in Figure~\ref{f:test_a_and_b}.

@@

To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
\todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance in the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
%
This tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment.
%
Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.

@@

% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments. 
%
For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish situations such as potentially dangerous close passing of a cyclist. A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives. 
% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
Therefore, for this case study, a threshold of $1$cm has been selected. Thus any variation of less than $1$cm is tolerable.

@@

\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$, 
where any value higher than the specified tolerance is considered non-permissible. }

@@

% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective. 
%

@@

\subsubsection{NUMA Control}
Control over Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multiprocessors with shared memory. This control allows the simulator program to be fixed on a single core reducing memory access time.
%
Initial screening tests done with NUMA control gave moderate improvements in simulation variance of the order of a few percent, see Section~\ref{s:screening}. This control may assist if simulation variance is borderline to the deterministic threshold but not seen as essential.

>>>>

\documentclass[letterpaper, 10 pt, journal, twoside]{IEEEtran}
% \input{archive/main.config.tex}
\usepackage[maxnames=6,firstinits=true,doi=false,url=false,isbn=false]{biblatex}
% \addbibresource{IROS.bib}
\addbibresource{cav_determinism.bib}
\usepackage{array}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{threeparttable}
\usepackage{balance} %balance columns
\hyphenation{op-tical net-works semi-conduc-tor}
\usepackage{enumitem}
\usepackage{hyperref} % Add  [ocgcolorlinks,pdfusetitle] before hyperref for colored links
% \usepackage{subfig}
\usepackage{subcaption}
\usepackage{dblfloatfix}
\usepackage{booktabs}
\usepackage{authblk}
\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
% \pagenumbering{gobble} %suppresses page numbers
\usepackage{balance} %balance columns
\usepackage{todonotes}
% \usepackage[disable]{todonotes}
\usepackage{siunitx}
% \usepackage{draftwatermark}

@@

\begin{abstract}
\noindent 
Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
%
Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
\todo[inline,color=green!40]{GC Alternative: Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
%
% KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
% These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
%
However, game engines do not need to be deterministic per se. % when used for games. 
%
This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation precision} for specific configurations and utilisation of the software and hardware, respectively. 
%
\todo[inline,color=red!40]{old: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup. }
A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine (game engine), illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.

\todo[inline,color=green!40]{GC Alternative: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance. }
\todo[inline,color=yellow!40]{AG Alternative: A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine (game engine), illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance. }
\todo[inline]{Review again at end. Potentially give away more on results to motivate the reader to continue.}

@@

% Game engines for SBV
Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation. 
%
However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include determinism.
%
% What is a deterministic system
In this context, deterministic refers to the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
% 
% Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions. Specifically, we focus on the variance of actor paths within a given scenario. 
%
Herein we adopt the engineering terms of \textit{precision} and \textit{tolerance} to describe how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions~\cite{ADictionaryofMechanicalEngineering}. %
%
As an analogy, the simulator can be thought of as manufacturing a product, producing data traces and if tasked to repeatedly make the same product we must ask how \textit{precise} can it be and what will the \textit{tolerance} be when comparing any differences between products. The precision of any computer based task is casually assumed to be absolute, but technically this is equivalent to the representation of numbers within it, e.g. 64-bit floating point \textbf{[ref]}. %
%
Real manufacturing can only aspire to achieve such precision and hence the need for tolerances on design specifications to account for the variance in real-world processes. If a simulator is deterministic then it will produce results with some degree of floating point precision and zero tolerance, i.e. with zero variance on any measure of distinction between simulation output traces. 
%
This is not to be confused with \textit{accuracy} which describes the extent to which the simulation complies to a known standard, in many cases a simulation is held to the standard of the real world, where any divergence from this standard is termed the \textit{reality gap}.

@@

% Verification and Determinism 
% Why is determinism important for verification
%For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
% KE Text
\todo[inline,color=red!40]{old: Determinism is a key prerequisite for simulation during AV development and testing. A deterministic simulation environment guarantees that tests are repeatable. Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated. This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.  }
%
Determinism is a key prerequisite for simulation during AV development and testing. A deterministic simulation environment guarantees that tests are repeatable. In conventional engineering terms a deterministic simulator can be considered to have zero \textit{tolerance}. If the tolerance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}. Therefore, \textit{tolerance} is the acceptable degree of variability between repeated execution outputs. When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated. This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.

@@

% Challenges of SBV
When used for gaming, game engines do not need to be deterministic nor do they even have any requirements on the limits of permissible tolerance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers. It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience. However, the situation is very different for AV development and testing. Thus, our main research questions is:
\todo[inline,color=red!40]{old: How can one assess the extent to which a simulation environment is deterministic?  }
How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?
\todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}

In pursuit of an answer to this question we have investigated how non-determinism of a game engine, namely Unreal Engine that powers CARLA, when used for simulation-based AV verification affects the simulation results. 
%
In our case study, scenarios between pedestrian and vehicle actors have been analysed to identify conditions that result in non-deterministic simulation traces by analysis of actor position variance.
%
\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified. }
%
By analysing actor position variance we find that the Carla simulator is not deterministic under certain conditions. However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
%
\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is not deterministic under certain conditions. However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}

The insights gained from this case study have motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment. 
%
\todo[inline,color=red!40]{old: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have. In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.  }
%
Knowing the simulation variance will help verifcation engineers assess the impact that using a game engine for AV simulation may have. In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
%
\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verifcation engineers assess the impact that using a game engine for AV simulation may have. In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this. }
%
% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
%We present a case study to illustrate the use of this method. The study highlights the effects of non-determinism of a game engine on simulation results. Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
%
%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
%

@@

Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.  %However, when these engines are used for AV development then deterministic behaviour is required. 
%
Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
%
\todo[inline]{GC line added to introduction. We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}. Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion. But it is important that we demonstrate knowledge of others working on this in other areas. {\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
%
\todo[inline,color=green!40]{GC Addition: Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}. }
%
This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
%
% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in AV simulations. Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.

@@

% This process allows hardware scalability, but the rendering will become of jerky quality on slower machines. These engines update at fixed intervals and render whenever they can, which is not steady. This results in what is so called residual lag\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, where the engine is trying to render between two consecutive updates. In this case, the engine will use extrapolation techniques to give an estimate of where the object should be. This often is sufficient for gaming purposes and unnoticeable to the user, in fact it improves the stuttery motion.
%
\todo[inline,color=red!40]{old: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences. Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy. In contrast, simulation accuracy is essential in AV development and testing, especially for perception stack testing, but also for visual inspection of accidents or near misses. If necessary, it is acceptable to achieve the required level of accuracy at the cost of real-time performance. For example, for perception stack testing the sensors need to get input that is as close as possible to what the real input would be, i.e.\ with no or minimal extrapolation or lag. This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.  }
%
Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences. Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision. The gamer neither wishes a faithful representation of reality, i.e. can be low accuracy, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e. can be low precision. In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses. If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance. For example, for perception stack testing the sensors need to get input that repeatable so that if any software bugs are found they can be re-played and the issue resolved. This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
\todo[inline,color=green!40]{GC Edit: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences. Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision. The gamer neither wishes a faithful representation of reality, i.e. can be low accuracy, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e. can be low precision. In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses. If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance. For example, for perception stack testing the sensors need to get input that repeatable so that if any software bugs are found they can be re-played and the issue resolved. This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.}
%
\todo[inline]{KIE: ... by slowing down the simulation to enable more extensive physics calculations?}

@@

Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
Thus, the results of arithmetic operations must fit into the given bit width, which is achieved through rounding to the nearest representable value. This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}. %
%
% SEE BELOW \todo[inline]{Check the points on floating point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour. Calculation results may differ due to the nature of non-associative floating point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}. Some authors even suggest avoiding the use of floating point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}. 
%
If calculations are perofrmed on a single core, then concurrency of multiple threads is not an issue.
\todo[inline,color=green!40]{GC Addition (added reference on non-associative fp addition Kapre2007, Whitehead2011, empirical-analysis-of-flaky-tests, see para below:)

Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour. Calculation results may differ due to the nature of non-associative floating point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}. Some authors even suggest avoiding the use of floating point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}. 
%
If calculations are perofrmed on a single core, then concurrency of multiple threads is not an issue.
% which is discussed in the section below titled Scheduling, Concurrency and Parallelisation.}
}
%
% Floating-point representation of real numbers is limited with fixed register sizes resulting in errors of rounding where loss of bits occur which are rounded or truncated and cancellation which happens where two similar values are subtracted.

@@

However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events. This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
%
Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
%
\todo[inline,color=green!40]{GC addition: Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}. }
%
Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.

@@

Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
%
\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}. }

@@

\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.

@@

\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
%
\todo[inline]{Complete the section structure as this is a very long section.}

@@

This case study draws on a setup used to verify an urban last-mile mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
%
Thus, the scene for our investigation is the T-junction and the scenarios are shown in Figure~\ref{f:test_a_and_b}.

@@

To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
%
Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance?
%
\todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
%
This tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment.
%
Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.

@@

% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments. 
%
For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish situations such as potentially dangerous close passing of a cyclist. A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives. 
% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
Therefore, for this case study and more broadly our general verifcation requirements, a threshold of $1$cm has been selected. Thus any variation of less than $1$cm is tolerable.

@@

@@

\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.  }

% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
%
\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$, 
where any value higher than the specified tolerance is considered non-permissible. }

@@

\subsubsection{NUMA Control}
Control over Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multiprocessors with shared memory. This control allows the simulator program to be fixed on a single core reducing memory access time.
%
Initial screening tests done with NUMA control gave moderate improvements in simulation variance of the order of a few percent, see Section~\ref{s:screening}. This control may assist if simulation variance is borderline to the permissible tolerance but not seen as essential.
